Face Recognition: Too Bias, or Not Too Bias?

### 2020-04-20 sections/analysis.tex
updated bar plot (Fig 6)
cleanup; removed -vspaces
wip

\section{Results and Analysis}
A single \gls{cnn} was used as a means to control the experiments. For this, Sphereface~\cite{liu2017sphereface} trained on CASIA-Web~\cite{yi2014learning}, and evaluated on \gls{lfw}~\cite{LFWTech} (\%99.22 accuracy), encoded all of the faces.\footnote{\href{$https://github.com/clcarwin/sphereface\_pytorch$}{https://github.com/clcarwin/sphereface\_pytorch}} As reported in~\cite{wang2019racial}, \gls{lfw} has about a 13\%, 14\%, 3\%, and 70\% ratio in Asians, Blacks, Indians, and Whites, respectively; furthermore, CASIA-Web is even more unbalanced (again, as reported in~\cite{wang2019racial}), with about  3\%, 11\%, 2\%, and 85\% for the same subgroups.
\vspace{-4mm}
\glsunset{det}
% \subsection{\gls{det} analysis}
\noindent\paragraph{\gls{det} analysis.}
\gls{det} curves (5-fold, averaged) show per-subgroup trade-offs (Fig.~\ref{fig:detcurves}). Note that \gls{m} performs better than \gls{f}, precisely as one would expect from the tails of score-distributions for \emph{genuine} pairs (Fig.~\ref{fig:detection-model}). \Gls{af} and \gls{if} perform the worst.
\glsunset{m}
\glsunset{f}

% Specifically, we analyze three \glspl{cnn} variants: we encode faces using VGG-16~\cite{simonyan2014very},  50-layer residual network (\ie ResNet-50)~\cite{he2016deep}, and \gls{senet} (\ie SENet-50)~\cite{hu2018squeeze}, with results of only the latter (\ie the best performing \gls{senet}) model used throughout the main paper. Results for the other two, alongside \gls{senet} for comparison, are in the Supplemental Material.
\vspace{-4mm}

\glsunset{am}\glsunset{af}\glsunset{bm}\glsunset{bf}\glsunset{im}\glsunset{if}\glsunset{wm}\glsunset{wf}
\begin{figure*}[h!]
    \centering
    \begin{subfigure}[t]{.3\linewidth}
    \includegraphics[width=.71\linewidth]{figures/curve_vgg16_subgroups-crop.pdf}
    \caption{VGG16~\cite{simonyan2014very}}
 \end{subfigure}
    \begin{subfigure}[t]{.27\linewidth}
    \includegraphics[width=.75\linewidth]{figures/curve_resnet50_subgroups-crop.pdf}
    \caption{ResNet50~\cite{he2016deep}}
   \end{subfigure}
    \begin{subfigure}[t]{.27\linewidth}
    \includegraphics[width=.75\linewidth]{figures/curve_senet50_subgroups-crop.pdf}
    \caption{SENet~\cite{hu2018squeeze}}
    \end{subfigure}
    \caption{\small{\textbf{\gls{det} curves for different CNNs}. \gls{fnr} (\%) (vertical) vs \gls{fpr}  (horizontal, log-scale) for VGG2~\cite{Cao18} models with different backbones (VGG16, Resnet50, SEnet50). Lower is better. For each plot, \gls{wm} is the top-performer, while \gls{af} is the worst. The the ordering of the curves is roughly the same for each backbone.}}\label{fig:sdm-appendix-a}
    \vspace{-5mm}
\end{figure*}
\noindent\paragraph{Score analysis.}
% \subsection{Score analysis}
Fig.~\ref{fig:detection-model} shows score distributions for faces of the same (\ie \emph{Genuine}) and different (\ie \emph{Imposter}) identity, with a subgroup per \gls{sdm} plot. Notice that score distributions for imposters tend to peak about zero for all subgroups, and with minimal deviation comparing modes of the different plots. On the other hand, the score distribution of the \emph{genuine} pairs varies across subgroups in location (\ie score value) and spread (\ie overall shape). Asian Fem
Fig.~\ref{fig:confusion} shows the confusion matrix of the subgroups. A vast majority of errors occur in the intra-subgroup. It is interesting to note that while the definition of  each group  based on ethnicity and race may not be crisply defined, the confusion matrix indicates that in practice the \gls{cnn} finds that the groups are effectively separate. The categories are, therefore, meaningful in for \gls{fr}.

@@

\begin{table}[b!]
    \vspace{-4mm}
\glsunset{tar}
\glsunset{far}
\caption{\small{\textbf{\gls{tar} at intervals of \gls{far}}. gls{far}, listed are the \gls{tar} scores for a global threshold (top) and the proposed category-based threshold (bottom). Higher is better.}}\label{tab:ethnicy-far} 
 \vspace{-2mm}
\begin{center}
\tiny
\begin{tabular}{l c c c c c}
     \gls{far} & 0.3 & 0.1 & 0.01 & 0.001 & 0.0001\\\midrule
    \multirow{2}{.1mm}{\textbf{\gls{af}}} &0.990 & 0.867 & 0.516 & 0.470 & 0.465\\[-4pt]
        &1.000 & 0.882 & 0.524 & 0.478 & 0.474\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{am}}} &0.994 & 0.883 & 0.529 & 0.482 & 0.477\\[-4pt]
        &1.000 & 0.890 & 0.533 & 0.486 & 0.482\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{bf}}} &0.991 & 0.870 & 0.524 & 0.479 & 0.473\\[-4pt]
        &1.000 & 0.879 & 0.530 & 0.484 & 0.480\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{bm}}} &0.992 & 0.881 & 0.526 & 0.480 & 0.474\\[-4pt]
        &1.000 & 0.891 & 0.532 & 0.485 & 0.480\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{if}}} &0.996 & 0.881 & 0.532 & 0.486 & 0.481\\[-4pt]
        &1.000 & 0.884 & 0.534 & 0.488 & 0.484\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{im}}} &0.997 & 0.895 & 0.533 & 0.485 & 0.479\\[-4pt]
        &1.000 & 0.898 & 0.535 & 0.486 & 0.481\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{wf}}} &0.988 & 0.878 & 0.517 & 0.469 & 0.464\\[-4pt]
        &1.000 & 0.894 & 0.526 & 0.478 & 0.474\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{wm}}} &0.989 & 0.896 & 0.527 & 0.476 & 0.470\\[-4pt]
        &1.000 & 0.910 & 0.535 & 0.483 & 0.478\\[-1pt]
    \midrule
    \multirow{2}{3mm}{\textbf{Avg.}} &0.992 & 0.881 & 0.526 & 0.478 & 0.474\\[-4pt]
        &1.000 & 0.891 & 0.531 & 0.483 & 0.479\\[-10pt]
\end{tabular}
\end{center}
\glsreset{tar}
\glsreset{far}
%  \vspace{-3mm}

@@

To further demonstrate the extend of the problem, we follow settings typical for systems in practice: we set the desired \gls{fpr}, and then determine the percent difference, \ie desired versus actual (Fig.~\ref{fig:percent:difference}, \emph{top}). Furthermore, we mitigate this highly skewed phenomena by applying subgroup-specific thresholds (\emph{bottom}). By this, minimal error from the desired \gls{fpr}. In addition, where there is small error, the offset is balanced across subgroups.

% \subsubsection{Other \gls{cnn} models.}~\label{app:sec:other:models}
\vspace{-5mm}
\noindent\paragraph{Model analysis.}
Variations in optimal threshold exist across models (Fig.~\ref{fig:sdm-appendix-a}). Like in Fig.~\ref{fig:detcurves}, the \gls{det} curves for three \gls{cnn}-based models, each trained on VGG2 with softmax but with different backbones.\footnote{\href{https://github.com/rcmalli/keras-vggface}{https://github.com/rcmalli/keras-vggface}} Notice similar trends across subgroups and models, which is consistent with  Sphereface as well (Fig.~\ref{fig:detcurves}). For example, the plots generated with Spherface and VggFace2 all have the \gls{wm} curve at the bottom (\ie best) and \gls{af} on top (\ie worst). Thus, the additional \gls{cnn}-based models demonstrate the same phenomena: proportional to the overall model performance, exact in the ordering of curves for each subgroup.

\vspace{-5mm}
\noindent\paragraph{Verification threshold analysis.}%\subsection{Verification threshold} \label{subsec:analysis:verification}
We seek to reduce the bias between subgroups. Such that an operating point (\ie \gls{fpr}) is constant across subgroups. To accomplish that, we used a per subgroup threshold. In \gls{fv}, we consider one image as the query and all others as the test. For this, the ethnicity of the query image is assumed. We can then examine the \gls{det} curves and pick the best threshold per subgroup for a certain \gls{fpr}.

@@

        &WF & 44.7 &43.8 & \textbf{\textcolor{blue}{57.3}}& \textbf{\textcolor{red}{48.0}} & \textbf{48.5} \\
        &WM & 30.1&\textbf{\textcolor{red}{47.4}} &  45.3 & \textbf{\textcolor{blue}{56.1}} & 44.7\\\cline{3-7}
        &Avg &  43.3& 47.4&\textbf{47.7} &46.3 &46.2\\
 \end{tabular}
 \end{tabular}
 }
 \end{center}
 \vspace{-7mm}
\end{table}

\vspace{-5mm}
\noindent\paragraph{Human evaluation analysis.}
% \subsection{Human evaluation}
Subjects of a subgroup likely have mostly been exposed to others of the same (Table~\ref{tab:humsn-eval-results} and Fig.~\ref{fig:human-eval}). Therefore, it is expected they would be best at labeling their own; similar for the same ethnicity, but another gender. Our findings concur. Each subgroup is best at labeling their type, and then second best at labeling the same ethnicity but opposite sex. Interestingly, each group of images is best tagged by the corresponding subgroup, with the second-best having the same ethnicity and opposite gender. On average, subgroups are comparable at labeling images. Analogous to the \gls{fr} system, performance ratings differ when analyzing within and across subgroups. In other words, performance on \gls{bfw} improved with subgroup-specific thresholds. Similarly, humans tend to better recognize individuals by facial cues of the same or similar demographics. Put differently, as the recognition performances drop with a global threshold optimized for one subgroup and deployed on another, human performance tends to drop when across subgroups (\ie performances drop for less familiar subgroups).

>>>>

\section{Results and Analysis}
A single \gls{cnn} was used as a means to control the experiments. For this, Sphereface~\cite{liu2017sphereface} trained on CASIA-Web~\cite{yi2014learning}, and evaluated on \gls{lfw}~\cite{LFWTech} (\%99.22 accuracy), encoded all of the faces.\footnote{\href{$https://github.com/clcarwin/sphereface\_pytorch$}{https://github.com/clcarwin/sphereface\_pytorch}} As reported in~\cite{wang2019racial}, \gls{lfw} has about a 13\%, 14\%, 3\%, and 70\% ratio in Asian, Black, Indian, and White, respectively. Furthermore, CASIA-Web is even more unbalanced (again, as reported in~\cite{wang2019racial}), with about  3\%, 11\%, 2\%, and 85\% for the same subgroups.
%\vspace{-4mm}
\glsunset{det}
% \subsection{\gls{det} analysis}
\noindent\paragraph{\gls{det} analysis.}
\gls{det} curves (5-fold, averaged) show per-subgroup trade-offs (Fig.~\ref{fig:detcurves}). Note that \gls{m} performs better than \gls{f}, precisely as one would expect from the tails of score-distributions for \emph{genuine} pairs (Fig.~\ref{fig:detection-model}). \Gls{af} and \gls{if} perform the worst.
\glsunset{m}
\glsunset{f}

% Specifically, we analyze three \glspl{cnn} variants: we encode faces using VGG-16~\cite{simonyan2014very},  50-layer residual network (\ie ResNet-50)~\cite{he2016deep}, and \gls{senet} (\ie SENet-50)~\cite{hu2018squeeze}, with results of only the latter (\ie the best performing \gls{senet}) model used throughout the main paper. Results for the other two, alongside \gls{senet} for comparison, are in the Supplemental Material.
%\vspace{-4mm}

\glsunset{am}\glsunset{af}\glsunset{bm}\glsunset{bf}\glsunset{im}\glsunset{if}\glsunset{wm}\glsunset{wf}
\begin{figure*}[h!]
    \centering
    \begin{subfigure}[t]{.3\linewidth}
    \includegraphics[width=.71\linewidth]{figures/curve_vgg16_subgroups-crop.pdf}
    \caption{VGG16~\cite{simonyan2014very}}
 \end{subfigure}
    \begin{subfigure}[t]{.27\linewidth}
    \includegraphics[width=.75\linewidth]{figures/curve_resnet50_subgroups-crop.pdf}
    \caption{ResNet50~\cite{he2016deep}}
   \end{subfigure}
    \begin{subfigure}[t]{.27\linewidth}
    \includegraphics[width=.75\linewidth]{figures/curve_senet50_subgroups-crop.pdf}
    \caption{SENet~\cite{hu2018squeeze}}
    \end{subfigure}
    \caption{\small{\textbf{\gls{det} curves for different CNNs}. \gls{fnr} (\%) (vertical) vs \gls{fpr}  (horizontal, log-scale) for VGG2~\cite{Cao18} models with different backbones (VGG16, Resnet50, SEnet50). Lower is better. For each plot, \gls{wm} is the top-performer, while \gls{af} is the worst. The ordering of the curves is roughly the same for each backbone.}}\label{fig:sdm-appendix-a}
%    \vspace{-5mm}
\end{figure*}
\noindent\paragraph{Score analysis.}
% \subsection{Score analysis}
Fig.~\ref{fig:detection-model} shows score distributions for faces of the same (\ie \emph{Genuine}) and different (\ie \emph{Imposter}) identity, with a subgroup per \gls{sdm} plot. Notice that score distributions for imposters tend to peak about zero for all subgroups, and with minimal deviation comparing modes of the different plots. On the other hand, the score distribution of the \emph{genuine} pairs varies across subgroups in location (\ie score value) and spread (\ie overall shape). Asian Fem
Fig.~\ref{fig:confusion} shows the confusion matrix of the subgroups. A vast majority of errors occur in the intra-subgroup. It is interesting to note that while the definition of  each group  based on ethnicity and race may not be crisply defined, the confusion matrix indicates that in practice, the \gls{cnn} finds that the groups are effectively separate. The categories are, therefore, meaningful for \gls{fr}.

@@

\begin{table}[b!]
%    \vspace{-4mm}
\glsunset{tar}
\glsunset{far}
\caption{\small{\textbf{\gls{tar} at intervals of \gls{far}}. \gls{far}, listed are the \gls{tar} scores for a global threshold (top) and the proposed category-based threshold (bottom). Higher is better.}}\label{tab:ethnicy-far} 
% \vspace{-2mm}
\begin{center}
\scriptsize
\begin{tabular}{l c c c c c}
     \gls{far} & 0.3 & 0.1 & 0.01 & 0.001 & 0.0001\\\midrule
    \multirow{2}{.1mm}{\textbf{\gls{af}}} &0.990 & 0.867 & 0.516 & 0.470 & 0.465\\[-4pt]
        &1.000 & 0.882 & 0.524 & 0.478 & 0.474\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{am}}} &0.994 & 0.883 & 0.529 & 0.482 & 0.477\\[-4pt]
        &1.000 & 0.890 & 0.533 & 0.486 & 0.482\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{bf}}} &0.991 & 0.870 & 0.524 & 0.479 & 0.473\\[-4pt]
        &1.000 & 0.879 & 0.530 & 0.484 & 0.480\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{bm}}} &0.992 & 0.881 & 0.526 & 0.480 & 0.474\\[-4pt]
        &1.000 & 0.891 & 0.532 & 0.485 & 0.480\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{if}}} &0.996 & 0.881 & 0.532 & 0.486 & 0.481\\[-4pt]
        &1.000 & 0.884 & 0.534 & 0.488 & 0.484\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{im}}} &0.997 & 0.895 & 0.533 & 0.485 & 0.479\\[-4pt]
        &1.000 & 0.898 & 0.535 & 0.486 & 0.481\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{wf}}} &0.988 & 0.878 & 0.517 & 0.469 & 0.464\\[-4pt]
        &1.000 & 0.894 & 0.526 & 0.478 & 0.474\\[-1pt]
    \multirow{2}{3mm}{\textbf{\gls{wm}}} &0.989 & 0.896 & 0.527 & 0.476 & 0.470\\[-4pt]
        &1.000 & 0.910 & 0.535 & 0.483 & 0.478\\[-1pt]
    \midrule
    \multirow{2}{3mm}{\textbf{Avg.}} &0.992 & 0.881 & 0.526 & 0.478 & 0.474\\[-4pt]
        &1.000 & 0.891 & 0.531 & 0.483 & 0.479\\[-10pt]
\end{tabular}
\end{center}
\glsreset{tar}
\glsreset{far}
%  \vspace{-3mm}

@@

To further demonstrate the extent of the problem, we follow settings typical for systems in practice. We set the desired \gls{fpr}, and then determine the percent difference, \ie desired versus actual (Fig.~\ref{fig:percent:difference}, \emph{top}). Furthermore, we mitigate this highly skewed phenomena by applying subgroup-specific thresholds (\emph{bottom}). By this, minimal error from the desired \gls{fpr}. In addition, where there is small error, the offset is balanced across subgroups.

% \subsubsection{Other \gls{cnn} models.}~\label{app:sec:other:models}
%\vspace{-5mm}
\noindent\paragraph{Model analysis.}
Variations in optimal threshold exist across models (Fig.~\ref{fig:sdm-appendix-a}). Like in Fig.~\ref{fig:detcurves}, the \gls{det} curves for three \gls{cnn}-based models, each trained on VGG2 with softmax but with different backbones.\footnote{\href{https://github.com/rcmalli/keras-vggface}{https://github.com/rcmalli/keras-vggface}} Notice similar trends across subgroups and models, which is consistent with  Sphereface as well (Fig.~\ref{fig:detcurves}). For example, the plots generated with Sphereface and VggFace2 all have the \gls{wm} curve at the bottom (\ie best) and \gls{af} on top (\ie worst). Thus, the additional \gls{cnn}-based models demonstrate the same phenomena: proportional to the overall model performance, exact in the ordering of curves for each subgroup.

%\vspace{-5mm}
\noindent\paragraph{Verification threshold analysis.}%\subsection{Verification threshold} \label{subsec:analysis:verification}
We seek to reduce the bias between subgroups. Such that an operating point (\ie \gls{fpr}) is constant across subgroups. To accomplish that, we used a per subgroup threshold. In \gls{fv}, we consider one image as the query, and all others as the test. For this, the ethnicity of the query image is assumed. We can then examine the \gls{det} curves and pick the best threshold per subgroup for a certain \gls{fpr}.

@@

        &WF & 44.7 &43.8 & \textbf{\textcolor{blue}{57.3}}& \textbf{\textcolor{red}{48.0}} & \textbf{48.5} \\
        &WM & 30.1&\textbf{\textcolor{red}{47.4}} &  45.3 & \textbf{\textcolor{blue}{56.1}} & 44.7\\\cline{3-7}
        &Avg &  43.3& 47.4&\textbf{47.7} &46.3 &46.2\\
 \end{tabular}
 \end{tabular}
 }
 \end{center}
% \vspace{-7mm}
\end{table}

%\vspace{-5mm}
\noindent\paragraph{Human evaluation analysis.}
% \subsection{Human evaluation}
Subjects of a subgroup likely have mostly been exposed to others of the same (Table~\ref{tab:humsn-eval-results} and Fig.~\ref{fig:human-eval}). Therefore, it is expected they would be best at labeling their own; similar for the same ethnicity, but another gender. Our findings concur. Each subgroup is best at labeling their type, and then second best at labeling the same ethnicity but opposite sex. Interestingly, each group of images is best tagged by the corresponding subgroup, with the second-best having the same ethnicity and opposite gender. On average, subgroups are comparable at labeling images. Analogous to the \gls{fr} system, performance ratings differ when analyzing within and across subgroups. In other words, performance on \gls{bfw} improved with subgroup-specific thresholds. Similarly, humans tend to better recognize individuals by facial cues of the same or similar demographics. Put differently, as the recognition performances drop with a global threshold optimized for one subgroup and deployed on another, human performance tends to drop when across subgroups (\ie performances drop for less familiar subgroups).

### 2020-04-20 sections/introduction.tex
wip

\glsresetall
\section{Introduction}
    As more of society becomes integrated with \gls{ml}, bias, fairness, and the formalization of \gls{ml} standards attract more attention~\cite{10.1007/978-3-030-13469-3_68, anne2018women, wang2018racial}. Thus, an effect of the growing dependency on \gls{ml} are growing concerns for biased and unfair algorithms, \eg untrustworthy and racist \gls{fr} systems~\cite{england2019,snow2018}.

    Typically, \glspl{cnn} are trained on faces identified by a detection system. Specifically, for \gls{fr}, the goal is to encode faces in an N-dimensional space such that samples of the same identity are neighbors and those of different identity of further apart. Thus, the goal is to identify subjects with face encodings. We can deploy a \gls{cnn} to encode faces, and then compare via a similarity score: assume pairs with a high enough score are \emph{genuine}; else, an \emph{imposter}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/fig1-crop.pdf}
    \caption{\small{\textbf{Depiction of biometrics.} The \gls{sdm} (\emph{top-left}) shows the sensitivity related to a single threshold $t_g$. Translated to error rate (\emph{bottom-left}), a direct trade-off between FNP and FPR. In practice, this is entirely application dependent. Specifically, the specific chose in desired FPR (\%) (\emph{bottom-right}).}}
    \label{fig:biometrics}
\end{figure}
    % \begin{figure}[t!]
    %     \centering
    %     \includegraphics[width=.8\linewidth]{figures/montage.pdf}
    %     \caption{\small{\textbf{\gls{bfw}.} Average face of the different subsets: \emph{top-left}: the entire \gls{bfw}; \emph{top-row} per race;  \emph{left-column}: per gender. The others represent the ethnicity and gender of the race and gender, respectively. Table~\ref{tab:ethnic-splits} defines the acronyms of subgroups.}}
    %     \label{fig:avg-faces}
    % \end{figure}

    When comparing scores, typically, a fixed threshold sets the decision boundary (Fig.~\ref{fig:biometrics}). Thus, features of the same identity must satisfy a criterion via a single value~\cite{deng2019arcface, liu2017sphereface, wang2018additive, wang2018cosface}. However, we found that a single (\ie global) threshold is a crude measure that leads to skewed errors - the held-out set used to determine the threshold tends to share the same distribution with the test data, which favors specific demographics that are a majority. That skew, the difference in the performance of an algorithm of certain demographics, is our definition of bias. A key question is: \emph{is \gls{fr} too biased, or not?}

@@

        % Acc$@t_o$ & 0.942 &0.952 &0.968 &0.961 &0.960 &0.962 & 0.974 &0.982 & 0.963 $\pm$ 0.012\\
        \bottomrule
    \end{tabular}}
    % \glsunset{af}
    }
    % % \glsunset{af}
    % \glsunset{am}
    % \glsunset{bf}
    % \glsunset{bm}
    % \glsunset{if}
    % \glsunset{im}
    % \glsunset{wf}
    % \glsunset{wm}
    % \vspace{-12pt}
\end{table*}

@@

    % The result is various types of biases in FR systems in favor of or against particular demographics remain a question.

    For \textbf{(1)} we propose subgroup-specific (\ie optimal) thresholds; to address \textbf{(2)}, we introduce a new benchmark for measuring bias in \gls{fr}, \gls{bfw} (Table~\ref{tab:compared} \&~\ref{tab:ethnic-splits}). \gls{bfw} allows for a fair evaluation of \gls{fr} systems, while enabling demographic-specific ratings to be reported. We use \gls{bfw} to gain a deeper understanding of the extent of bias present in \gls{soa} \gls{cnn} used to encode faces. Then, we suggest a mechanism to mitigate problems of bias with more balanced performance ratings for different demographics. Specifically, we propose using an adaptive threshold that varies depending on the characteristics of detected facial attributes (\ie gender and ethnicity, Fig~\ref{fig:avg-faces}). We show an increase in accuracy with a balanced performance for different subgroups. Similarly, we show a positive effect of adjusting the similarity threshold based on the facial features of matched faces. Thus, selective use of similarity thresholds in current \gls{soa} \gls{fr} systems provides more intuition in \gls{fr} research with a method easy to adopt in practice.

    The contributions of this work are 3-fold. (1) We built a balanced dataset as a proxy to measure verification performance per subgroup for studies of bias in \gls{fr}. (2) We revealed an unwanted bias in scores of face pairs - a bias that causes ratings to skew across demographics. For this, we showed that an adaptive threshold per subgroup balances performance (\ie the typical use of a global threshold unfavorable, which we address via optimal thresholds). (3) We surveyed humans to demonstrate bias in human perception (NIH-certified, \textit{Protect Humans in Research}).%\footnote{NIH-certified, \textit{Protect Humans in Research}.}%19-09-08}.}
    % \begin{enumerate}
    %     \item Built a balanced dataset as a proxy to measure verification performance per subgroup.
    %     \item Analyzed an unwanted bias in scores of face pairs while showing that optimal thresholds determined per subgroup significantly boost and balance performances.
    %     \item Showed bias causes inconsistencies in ratings across demographics-- the typical use of a global threshold unfavorable. We mitigate the problem via adaptive thresholds.
    %     \item Conducted human-evaluations to demonstrate bias in human perception.\footnote{NIH-certified, \textit{Protect Humans in Research}, \textcolor{red}{IRB 19-09-08}.}
    % \end{enumerate}

>>>>

\glsresetall
\section{Introduction}
    As more of society becomes integrated with \gls{ml}, topics related to bias, fairness, and the formalization of \gls{ml} standards attract more attention~\cite{10.1007/978-3-030-13469-3_68, anne2018women, wang2018racial}. Thus, an effect of the growing dependency on \gls{ml} is a growing concern for biased and unfair algorithms, \eg untrustworthy and racist \gls{fr} systems~\cite{england2019, nagpal2019deep,snow2018}.

    Nowadays, \glspl{cnn} are trained on faces identified by a detection system. Specifically, for \gls{fr}, the goal is to encode faces in an N-dimensional space such that samples of the same identity are neighbors, while different identities are farther apart. Thus, we can deploy a \gls{cnn} to encode faces, and then compare via a similarity score (\ie pairs with a high enough score classified as \emph{genuine}; else, \emph{imposter}).

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/fig1-crop.pdf}
    \caption{\small{\textbf{Depiction of bio-metrics.} The \gls{sdm} (\emph{top-left}) shows the sensitivity related to a single threshold $t_g$. Translated to error rate (\emph{bottom-left}), a direct trade-off between FNP and FPR. In practice, this is entirely application dependent. Specifically, the specific chose in desired FPR (\%) (\emph{bottom-right}).}}
    \label{fig:biometrics}
\end{figure}

Typically, a fixed threshold sets the decision boundary by which to compare scores (Fig.~\ref{fig:bio-metrics}). Thus, features of the same identity must satisfy a criterion based on a single value~\cite{deng2019arcface, liu2017sphereface, wang2018additive, wang2018cosface}. However, we found that a single (\ie global) threshold is a crude measure that leads to skewed errors. Furthermore, the held-out set used to determine the threshold tends to share the same distribution with the test set, favoring specific demographics that are the majority. That skew (\ie the difference in the performance of an algorithm of certain demographics) is our definition of bias. A key question is: \emph{is \gls{fr} too biased, or not?}

@@

        % Acc$@t_o$ & 0.942 &0.952 &0.968 &0.961 &0.960 &0.962 & 0.974 &0.982 & 0.963 $\pm$ 0.012\\
        \bottomrule
    \end{tabular}}
    }
    % \vspace{-12pt}
\end{table*}

@@

    For \textbf{(1)} we propose subgroup-specific (\ie optimal) thresholds, while we address \textbf{(2)} with a new benchmark dataset for measuring bias in \gls{fr}, \gls{bfw} (Table~\ref{tab:ethnic-splits} and~\ref{tab:compared}). \gls{bfw} allows for a fair evaluation of \gls{fr} systems, while enabling demographic-specific ratings to be reported. We use \gls{bfw} to gain a deeper understanding of the extent to which bias presents in \gls{soa} \gls{cnn} used to encode faces. Then, we suggest a mechanism to mitigate problems of bias with more balanced performance ratings for different demographics. Specifically, we propose using an adaptive threshold that varies depending on the characteristics of detected facial attributes (\ie gender and ethnicity, Fig~\ref{fig:avg-faces}). We show an increase in accuracy with a balanced performance for different subgroups. Similarly, we show a positive effect of adjusting the similarity threshold based on the facial features of matched faces. Thus, selective use of similarity thresholds in current \gls{soa} \gls{fr} systems provides more intuition in \gls{fr} research with a method easy to adopt in practice.

    The contributions of this work are 3-fold. (1) We built a balanced dataset as a proxy to measure verification performance per subgroup for studies of bias in \gls{fr}. (2) We revealed an unwanted bias in scores of face pairs - a bias that causes ratings to skew across demographics. For this, we showed that an adaptive threshold per subgroup balances performance (\ie the typical use of a global threshold unfavorable, which we address via optimal thresholds). (3) We surveyed humans to demonstrate bias in human perception (NIH-certified, \textit{Protect Humans in Research}).%\footnote{NIH-certified, \textit{Protect Humans in Research}.}%19-09-08}.}
